---
title: "Hands-on Session: Multiple Ascending Dose PK/PD Binary Response Analysis"
subtitle: "Exploratory Analysis, Dose-Response, Longitudinal Summaries, and nlmixr2 Modeling"
author: "Teaching Session"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

# Learning objectives

By the end of this hands-on session, students should be able to:

- Understand the structure of a simulated multiple ascending dose PK/PD dataset.
- Filter and prepare binary pharmacodynamic response data.
- Summarise binary responder rates using exact binomial confidence intervals.
- Visualise longitudinal response patterns over time.
- Explore dose-response relationships.
- Prepare data for categorical modeling in `nlmixr2`.
- Fit a simple binary response model.
- Generate a categorical visual predictive check.

# Dataset description

This session uses a model-generated multiple ascending dose dataset that mimics an orally administered small molecule.

The dataset includes pharmacokinetic and pharmacodynamic endpoints:

| Column | Description |
|---|---|
| `ID` | Unique subject ID |
| `TIME` | Time relative to first drug administration |
| `NOMTIME` | Nominal time |
| `TIMEUNIT` | Unit of time |
| `AMT` | Dosing amount in mg |
| `LIDV` | Observation value |
| `CMT` | Compartment or endpoint identifier |
| `NAME` | Description of event |
| `EVENTU` | Unit of observation |
| `CENS` | Censoring indicator |
| `EVID` | Event ID: 0 observation, 1 dosing |
| `WEIGHTB` | Baseline body weight |
| `SEX` | Sex |
| `TRTACT` | Treatment group label |
| `DOSE` | Randomised dose in mg |
| `PROFDAY` | Profile day |
| `PROFTIME` | Time within profile day |
| `CYCLE` | Number of drug administrations received |

Compartment mapping:

| `CMT` | Meaning |
|---|---|
| 1 | Dosing event |
| 2 | PK concentration |
| 3 | Continuous response |
| 4 | Count response |
| 5 | Ordinal response |
| 6 | Binary response |

# Setup

```{r}
library(gridExtra)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(xgxr)
library(RColorBrewer)
library(binom)
library(nlmixr2)
library(rxode2)
library(MASS)
```

```{r}
set.seed(12345)

status <- "DRAFT"

xgx_theme_set()
```

# Define analysis constants

```{r}
DOSE_CMT <- 1
PK_CMT <- 2
PD_CMT <- 6

SS_PROFDAY <- 6
PD_PROFDAYS <- c(0, 2, 4, 6)

time_units_dataset <- "hours"
time_units_plot <- "days"

trtact_label <- "Dose"
dose_label <- "Dose (mg)"
pd_label <- "Responder Rate (%)"
```

# Load data

The `mad` dataset is available from the `xgxr` package.

```{r}
data("mad", package = "xgxr")

data <- mad
```

```{r}
glimpse(data)
```

# Prepare dataset

We create:

- `PROFDAY`: profile day derived from nominal time.
- `DAY_label`: readable day label.
- `TRTACT_low2high`: treatment factor ordered from low to high dose.
- `TRTACT_high2low`: treatment factor ordered from high to low dose.

```{r}
data <- data |>
  mutate(
    PROFDAY = case_when(
      NOMTIME < (SS_PROFDAY - 1) * 24 ~ 1 + floor(NOMTIME / 24),
      TRUE ~ SS_PROFDAY
    ),
    DAY_label = ifelse(PROFDAY > 0, paste("Day", PROFDAY), "Baseline")
  ) |>
  arrange(DOSE) |>
  mutate(
    TRTACT_low2high = factor(TRTACT, levels = unique(TRTACT)),
    TRTACT_high2low = factor(TRTACT, levels = rev(unique(TRTACT)))
  )
```

```{r}
glimpse(data)
```

# Prepare binary PD data

For this session, we focus on the binary pharmacodynamic endpoint.

```{r}
pd_data <- data |>
  filter(CMT == PD_CMT) |>
  mutate(
    LIDV_jitter = jitter(LIDV, amount = 0.1),
    DAY_jitter = jitter(PROFDAY, amount = 0.1)
  )
```

```{r}
glimpse(pd_data)
```

# Helper functions

## Binomial confidence interval

```{r}
binom_ci <- function(x, n) {
  ci <- binom::binom.exact(x, n)
  list(lower = ci$lower, upper = ci$upper)
}
```

## Summarise binary response

```{r}
summarise_binom <- function(df, group_vars) {
  df |>
    group_by(across(all_of(group_vars))) |>
    summarise(
      n = n(),
      responders = sum(LIDV, na.rm = TRUE),
      prop = responders / n,
      lower = binom::binom.exact(responders, n)$lower,
      upper = binom::binom.exact(responders, n)$upper,
      .groups = "drop"
    )
}
```

## Time and response scales

```{r}
x_time_scale <- function() {
  xgx_scale_x_time_units(
    units_dataset = time_units_dataset,
    units_plot = time_units_plot
  )
}

y_resp_scale <- function(limits = c(0, 1)) {
  scale_y_continuous(labels = percent, limits = limits)
}
```

# Colour palette

```{r}
dose_levels_active <- pd_data |>
  filter(DOSE > 0) |>
  distinct(TRTACT_low2high) |>
  pull(TRTACT_low2high) |>
  as.character()

dose_pal <- colorRampPalette(brewer.pal(8, "Dark2"))(
  length(dose_levels_active)
)

names(dose_pal) <- dose_levels_active
```

# Exploratory analysis

## Binary responder rate over time by dose

```{r}
ggplot(pd_data, aes(x = NOMTIME, y = LIDV, color = TRTACT_high2low)) +
  xgx_stat_ci(
    distribution = "binomial",
    position = position_dodge(width = 12),
    alpha = 0.5
  ) +
  x_time_scale() +
  y_resp_scale(c(0, 1)) +
  labs(
    x = "Time (days)",
    y = pd_label,
    color = trtact_label,
    title = "Responder Rate Over Time by Dose"
  ) +
  xgx_annotate_status(status)
```

## Teaching question

- What trend do you observe over time?
- Are higher doses associated with higher responder rates?
- At which time points are dose differences most visible?

# Dose-response at steady state

We define Day 6 as the steady-state profile day.

```{r}
pd_data_ss <- pd_data |>
  filter(PROFDAY == SS_PROFDAY)
```

```{r}
gg_dose_ss <- ggplot(pd_data_ss, aes(x = DOSE, y = LIDV)) +
  xgx_stat_ci(
    conf_level = 0.95,
    distribution = "binomial",
    geom = list("point", "errorbar")
  ) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial(link = "logit")),
    color = "black"
  ) +
  y_resp_scale(c(0, 1)) +
  labs(
    x = dose_label,
    y = pd_label,
    title = "Dose-Response at Day 6"
  ) +
  xgx_annotate_status(status)
```

```{r}
grid.arrange(
  gg_dose_ss +
    ggtitle("Dose on Linear Scale"),
  gg_dose_ss +
    scale_x_log10(breaks = unique(pd_data_ss$DOSE)) +
    ggtitle("Dose on Log Scale"),
  nrow = 1
)
```

## Teaching question

- Does the response appear to plateau?
- Is the relationship approximately monotonic?

# Dose-response by selected profile days

```{r}
pd_data_days <- pd_data |>
  filter(PROFDAY %in% PD_PROFDAYS)
```

```{r}
ggplot(pd_data_days, aes(x = DOSE, y = LIDV)) +
  xgx_stat_ci(
    conf_level = 0.95,
    distribution = "binomial",
    geom = list("point", "errorbar")
  ) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial(link = "logit")),
    color = "black"
  ) +
  y_resp_scale(c(0, 1)) +
  labs(
    x = dose_label,
    y = pd_label,
    title = "Dose-Response by Profile Day"
  ) +
  facet_grid(~DAY_label) +
  xgx_annotate_status(status)
```

# Exposure-response exploration

Next, we join PK concentration and binary PD response records into a wide format.

```{r}
pkpd_data_wide <- data |>
  filter(CMT %in% c(PK_CMT, PD_CMT)) |>
  transmute(
    ID,
    PROFDAY,
    NOMTIME,
    CENS,
    TRTACT_low2high,
    TRTACT_high2low,
    endpoint = case_when(
      CMT == PK_CMT ~ "CONC",
      CMT == PD_CMT ~ "PD"
    ),
    value = LIDV
  ) |>
  tidyr::pivot_wider(
    names_from = endpoint,
    values_from = value
  ) |>
  filter(!is.na(CONC), !is.na(PD))
```

```{r}
glimpse(pkpd_data_wide)
```

```{r}
pd_binary_label <- "Binary response"
conc_label <- "PK concentration"
```

```{r}
gg <- ggplot(pkpd_data_wide, aes(y = CONC, x = factor(PD))) +
  geom_boxplot(
    aes(group = factor(PD)),
    width = 0.5,
    outlier.shape = NA,
    size = 0.7
  ) +
  geom_jitter(
    data = pkpd_data_wide |>
      filter(CENS == 0, PROFDAY == 5),
    aes(color = TRTACT_high2low),
    shape = 19,
    width = 0.1,
    height = 0,
    alpha = 0.5
  ) +
  geom_jitter(
    data = pkpd_data_wide |>
      filter(CENS == 1, PROFDAY == 5),
    color = "red",
    shape = 8,
    width = 0.1,
    height = 0,
    alpha = 0.5
  ) +
  xgx_scale_y_log10() +
  labs(
    x = pd_binary_label,
    y = conc_label,
    color = trtact_label,
    title = "Exposure by Binary Response"
  ) +
  coord_flip()

print(gg)
```

## Teaching question

- Do responders appear to have higher concentrations than non-responders?

# Longitudinal responder summaries

We summarise responder rates over time by active dose.

```{r}
pd_sum_active <- summarise_binom(
  pd_data |>
    filter(DOSE > 0),
  c("NOMTIME", "TRTACT_low2high", "TRTACT_high2low", "DOSE")
)
```

```{r}
glimpse(pd_sum_active)
```

# Dosing event overlays

```{r}
dose_events <- data |>
  filter(CMT == DOSE_CMT, DOSE > 0) |>
  distinct(TRTACT_low2high, DOSE, NOMTIME, PROFDAY) |>
  arrange(DOSE, NOMTIME)

dose_events_all <- data |>
  filter(CMT == DOSE_CMT, DOSE > 0) |>
  distinct(NOMTIME, PROFDAY) |>
  arrange(NOMTIME)
```


# Faceted longitudinal responder rate by dose

```{r}
gg_facet_with_dose <- ggplot(
  pd_sum_active,
  aes(
    x = NOMTIME,
    y = prop,
    color = TRTACT_low2high,
    fill = TRTACT_low2high
  )
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.22,
    color = NA
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_vline(
    data = dose_events,
    aes(xintercept = NOMTIME),
    linetype = "dashed",
    color = "gray40",
    alpha = 0.5,
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  x_time_scale() +
  y_resp_scale(c(0, 1)) +
  scale_color_manual(values = dose_pal, guide = "none") +
  scale_fill_manual(values = dose_pal, guide = "none") +
  labs(
    x = "Time (days)",
    y = pd_label,
    title = "Longitudinal Responder Rate Over Time Stratified by Dose",
    subtitle = "Placebo excluded; dashed lines show dosing events"
  ) +
  facet_wrap(~TRTACT_low2high) +
  xgx_annotate_status(status) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )

print(gg_facet_with_dose)
```

# Overall active treatment responder rate

```{r}
pd_sum_overall_active <- pd_sum_active |>
  group_by(NOMTIME) |>
  summarise(
    n = sum(n),
    responders = sum(responders),
    prop = responders / n,
    lower = binom::binom.exact(responders, n)$lower,
    upper = binom::binom.exact(responders, n)$upper,
    .groups = "drop"
  )
```

```{r}
gg_overall_with_dose <- ggplot(
  pd_sum_overall_active,
  aes(x = NOMTIME, y = prop)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    fill = "#80CDC1",
    alpha = 0.35
  ) +
  geom_line(
    color = "#018571",
    linewidth = 1.2
  ) +
  geom_point(
    color = "#01665E",
    size = 2.2
  ) +
  geom_vline(
    data = dose_events_all,
    aes(xintercept = NOMTIME),
    linetype = "dashed",
    color = "gray40",
    alpha = 0.6,
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  x_time_scale() +
  y_resp_scale(c(0, 1)) +
  labs(
    x = "Time (days)",
    y = pd_label,
    title = "Longitudinal Responder Rate Over Time Overall",
    subtitle = "Pooled active treatment arms; placebo excluded; dashed lines show dosing events"
  ) +
  xgx_annotate_status(status) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray30")
  )

print(gg_overall_with_dose)
```

# Study design summaries

```{r}
study_design_table <- data |>
  group_by(TRTACT_low2high, DOSE) |>
  summarise(
    n_patients = n_distinct(ID),
    n_total_records = n(),
    n_pd_records = sum(CMT == PD_CMT, na.rm = TRUE),
    n_dose_records = sum(CMT == DOSE_CMT, na.rm = TRUE),
    n_nominal_times = n_distinct(NOMTIME),
    nominal_times = paste(sort(unique(NOMTIME)), collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(DOSE)

print(study_design_table)
```


# Prepare data for nlmixr2

For the model, we keep dosing records and binary PD records.

```{r}
fit_data <- data |>
  mutate(
    ID = factor(ID),
    TIME = as.numeric(TIME),
    NOMTIME = as.numeric(NOMTIME),
    DV = as.numeric(LIDV),
    WEIGHTB = as.numeric(WEIGHTB),
    SEX = case_when(
      SEX %in% c("Male", "MALE", "male", "1", 1) ~ 1,
      TRUE ~ 0
    ),
    PROFDAY = as.numeric(PROFDAY),
    TRTACT = factor(TRTACT_low2high, levels = levels(TRTACT_low2high))
  ) |>
  filter(NAME %in% c("Dosing", "PD - Binary")) |>
  filter(TRTACT != "Placebo") |>
  arrange(ID, TIME)
```

```{r}
glimpse(fit_data)
```

# Binary response model using nlmixr2

This model describes the probability of binary response using a logit relationship.

The linear predictor is:

$$
\text{logit}(p_i) =
\theta_0 +
\text{treatment effect} +
\text{weight effect} +
\text{sex effect} +
\text{time effect}
$$

Then:

$$
Y_i \sim \text{Binomial}(n = 1, p_i)
$$

```{r}
categorical_nlmixr <- function() {
  ini({
    theta0_pop <- -3.18
    beta_time_pop <- -0.05

    beta_theta0_TRTACT_1600_mg <- 0.60
    beta_theta0_TRTACT_200_mg <- -0.48
    beta_theta0_TRTACT_400_mg <- -0.11
    beta_theta0_TRTACT_800_mg <- 0.67

    beta_theta0_WEIGHTB <- -0.01
    beta_theta0_SEX_Male <- 1.68
    beta_beta_time_SEX_Male <- -1.98

    eta.theta0 ~ 1.8e-06
    eta.beta_time ~ 0.139

    n <- fix(1)
  })

  model({
    trt_1600 <- (TRTACT == "1600 mg")
    trt_200  <- (TRTACT == "200 mg")
    trt_400  <- (TRTACT == "400 mg")
    trt_800  <- (TRTACT == "800 mg")

    theta0 <- theta0_pop +
      beta_theta0_TRTACT_1600_mg * trt_1600 +
      beta_theta0_TRTACT_200_mg * trt_200 +
      beta_theta0_TRTACT_400_mg * trt_400 +
      beta_theta0_TRTACT_800_mg * trt_800 +
      beta_theta0_WEIGHTB * WEIGHTB +
      beta_theta0_SEX_Male * SEX +
      eta.theta0

    beta_time <- exp(
      beta_time_pop +
        beta_beta_time_SEX_Male * SEX +
        eta.beta_time
    )

    lp <- theta0 + beta_time * PROFDAY

    p1 <- 1 / (1 + exp(-lp))

    var ~ binom(n, p1)
  })
}
```

# Fit model

Depending on the computer and package versions, this step may take some time.

```{r,eval=T}
fit_nlmixr2 <- nlmixr2(
  categorical_nlmixr,
  fit_data,
  est = "focei",
  control = foceiControl(outerOpt = "bobyqa")
)
```

```{r}
#fit_nlmixr2 = readRDS("fit_nlmixr2.RDS")
print(fit_nlmixr2)
```


# Optional categorical VPC

This section assumes that a helper function named `plot_categorical_vpc_nlmixr2()` is available.

```{r}
vpc_function_file <- "vpc_from_nlmixr2_fit.R"

if (file.exists(vpc_function_file)) {
  source(vpc_function_file)
} else {
  message("VPC helper function file was not found. Skipping VPC section.")
}
```

```{r}
if (exists("plot_categorical_vpc_nlmixr2")) {
  p_vpc_nlmixr2 <- plot_categorical_vpc_nlmixr2(
    fit = fit_nlmixr2,
    data = fit_data,
    id_col = "ID",
    time_col = "TIME",
    dv_col = "DV",
    pred_var = "p1",
    nBins = 10,
    nSim = 200,
    ci = 0.95,
    seed = 12345,
    x_transform = function(x) x / 24,
    xlab = "Time (days)",
    ylab = "P(Y = 1)",
    title = "Categorical VPC from nlmixr2 Fit"
  )

  print(p_vpc_nlmixr2)
}
```



# Key takeaways

- Binary pharmacodynamic data can be summarised using responder rates and binomial confidence intervals.
- Dose-response and exposure-response plots provide complementary insights.
- Longitudinal visualisation helps assess onset, accumulation, and potential plateauing of response.
- Dosing event overlays are useful for interpreting repeated-dose studies.
- Categorical models can describe binary response probabilities while accounting for covariates and interindividual variability.
